Gaps in stellar streams as a result of globular cluster fly-bys

### 2025-05-13 paper/main.tex
editing the figure to have the right dimensions
explaining experiment in the appendix

\textbf{\paola{In Section~\ref{sec:mass_loss}, we emphasized that the conditions chosen to simulate the formation and evolution of streams around Pal~5 are probably unrealistic. By choosing, as we did, the same values for the mass and size of the cluster 5 billion years ago as the cluster has today, we did not take into account the loss of mass and internal evolution of the cluster over time. In particular, the velocity dispersion of the stars in a cluster and the velocity dispersion of the stars leaving the cluster and entering the stream depend on the cluster's mass. It is therefore legitimate to ask whether, using a progenitor with a different mass and/or size, the gaps survive and whether, in particular, gaps formed several billion years ago can still be visible today. }}

\begin{figure*}
    \centering
    \includegraphics[width=1\linewidth]{gap_no_more.png}
    \caption{When the mass of Palomar~5 is increased, the gap is lost.}
    \label{fig:gap_no_more.png}
\end{figure*}

\textbf{\paola{While we anticipate that a specific study on this aspect is in preparation (Ferrone et al, in prep), we would like to preview a few results here.
Fig.~\ref{fig:gap_no_more} (top and bottom panels) shows, respectively, a stream generated in one of our simulations, using as initial mass (i.e., 5 billion years ago) for Pal~5 one similar to the current one or a larger mass ($xx M_\odot$). For simplicity, in both experiments, we left the cluster size unchanged. In both cases, a gap forms due to an encounter (specifically with the cluster NGCxxx) that occurred xx billion years ago. However, while in Fig.~\ref{fig:gap_no_more}, top panel, the gap persists for xx billion years, in Fig. xx (bottom panel), the gap is no longer observable at the end of the simulation. Looking more closely at the reasons for this difference, we see (Fig.\ref{fig:decoherence_shocked_outflow}) that the gap formed xx billion years ago in Fig. xx (bottom) survives in all the populations lost from the cluster before the impact and which had time to reach the part of the tail affected by the passage of NGCxxx. What happens, however, is that after a coherent gap has formed in the tail, and in particular in the populations of stars in this part of the tail that escaped from the cluster at different times, in each of these populations the respective gap tends to travel along the streams at different speeds, which leads to a phase shift of the gap and its loss of visibility. In other words, the gap can survive for billions of years in each population of stars that left the cluster at different times, but it is no longer. Still, initially coherent and therefore visible because each of these populations travels along the tail (and drags a gap with it) at different speeds. The loss of coherence occurs at a time $t_{visibility}$ which is directly proportional to the gap size and inversely proportional to the velocity dispersion in that part of the stream. So, if populations of stars lost at different times co-exist in the portion of the stream where the impact occurs, the gap will remain visible for a time that depends on the mass of the progenitor system at the time these stars escaped. From this, we can understand that various factors determine the visibility of a gap in a stream. Mass affects the velocity dispersion in a given region of the stream, but this velocity dispersion, in turn, depends on the history of mass loss of the cluster. Suppose stars dominate a given region of the stream, all lost simultaneously from the cluster. In that case, the resulting velocity dispersion is lower than if various populations coexist in the same region, which has implications for these structures' persistence. }}

\begin{figure*}
    \includegraphics[width=\linewidth]{decoherence_shocked_outflow.png}
    \caption{Nice}
    \label{fig:decoherence_shocked_outflow}
\end{figure*}

>>>>

In Section~\ref{sec:mass_loss}, we emphasized that the conditions chosen to simulate the formation and evolution of streams around Pal~5 are probably unrealistic. By choosing, as we did, the same values for the mass and size of the cluster 5 billion years ago as the cluster has today, we did not take into account the loss of mass and internal evolution of the cluster over time. In particular, the velocity dispersion of the stars in a cluster and the velocity dispersion of the stars leaving the cluster and entering the stream depend on the cluster's mass. It is therefore legitimate to ask whether, using a progenitor with a different mass and/or size, the gaps survive and whether, in particular, gaps formed several billion years ago can still be visible today.

\begin{figure*}
    \centering
    \includegraphics[width=1\linewidth]{gap_no_more.png}
    \caption{When the mass of Palomar~5 is increased, the gap is lost.}
    \label{fig:gap_no_more}
\end{figure*}

While we anticipate that a specific study on this aspect is in preparation (Ferrone et al, in prep), we would like to preview a few results here.
Fig.~\ref{fig:gap_no_more} shows two streams with the same orbital parameters that were impacted by NGC~7078. The top panel shows a stream that comes from a host mass of $10,000$~$M_\odot$, while in the second panel we increased the mass to $\sim40,000$~$M_\odot$. For simplicity, both clusters have the same half-mass radius of 27~pc. The encounter occurred about two and a half billion years ago by NGC~7078, which has a mass of XX. The gap persists for the lighter host, yet is barely visible in the case of the more massive host.

\salvatore{Fig.~\ref{fig:decoherence_shocked_outflow} dissects the more massive stream of Fig.~\ref{fig:gap_no_more} and instructs us why the gap does not persist. The bottom panel shows the total density distribution, and dissects it into two profiles: stars that either are \textit{leaked} or \textit{shocked} out of the cluster. By shocked, we mean the stars that are lost during pericenter passages. Tidal forces generally scale with $r^{-3}$, thus significantly strengthening through their pericenter passages. See Fig.~3 of \citet{2004AJ....127.2753D} for an illustration of the magnitude of the tidal forces along an orbit. A consequence of the pericenter passages is that many more stars escape, fleeing with a higher mean velocity and velocity dispersion, which can be visualized in the \textit{Christmas tree} illustration in Fig.~A.3 of \citet{2012A&A...546L...7M}. We consider stars that leave between pericenter passages to be \textit{leaked} stars. We identify pericenter passages by computing each local minimum of the cluster's galactic-centric 3D radius with \texttt{scipy.signal.find\_peaks}. The duration of each shock is: $t_{\textrm{shock}}$ = 2$\sqrt{r/|\nabla\Phi|}$, where $r$ is the galactocentric radius and $|\nabla\Phi|$ is the magnitude of the acceleration from the galactic potential. Stars within this range about a given pericenter passage are all given the same color and shown as individual groups in the first, second, and third panels of Fig.~\ref{fig:decoherence_shocked_outflow}.}

\salvatore{The situation can be simplified as one-dimensional collisionless dispersion. These packets of stars can be described as a population with the same initial position that drifts about a mean velocity, $\langle v \rangle$, with velocity dispersion, $\sigma_v$. The density evolution is a Gaussian whose amplitude decreases inversely with time, and whose standard deviation grows linearly with time:}
\begin{equation}    
        \rho(x,t) = \frac{1}{\sigma_v t \sqrt{2\pi} }\textrm{exp}\left(-\frac{1}{2}\left(\frac{x-\langle v \rangle t}{\sigma_v t}\right)^2\right).
\end{equation}

\salvatore{It is important to note that within a single packet, the stars self-segregate. The position of the density peak is at $\langle v \rangle t$, while the faster stars are ahead and the slower ones are behind. The packets are separated by a distance of $\langle  v\rangle T$, where $T$ is the orbital period. Regarding our simulations, since the mass and radius of the host are held constant, the mean velocity and velocity dispersion are the same across all packets. The density of each packet decreases since the easier-to-remove stars are stripped early on. }

\salvatore{Consider an impact that happens at $x\prime$. The distance from the impact site to each packet is: $\delta x= x\prime - \langle v\rangle \left(t-n T\right)$, where $n$ is the packet index, with $n=0$ being the first packet escaped at the first pericenter passage in the simulation. Note that $\delta x\propto n$, and velocity segregation means $v\propto x$. Therefore, the different parts of the separate packets are hit at a given impact site. The earlier packets are hit in regions where the stars have lower velocities than the later packets. As a result, the gaps in the later packets quickly overtake those from the earlier packets. As a result, the gaps across the packets go out of phase and interfere destructively.}

\salvatore{In the simplest consideration, of two gaps of the same dimension $\sigma_x$ and an offset of $\delta x$, the gaps are out of phase when $1<\sigma_x/\delta x$. In the temporal analysis of the stream presented in Fig.~\ref{fig:decoherence_shocked_outflow}, we did not observe at any point of the simulation that the gap initially persisted to then go out of phase. Drawing from the wisdom in \citet{2015MNRAS.450.1136E} \& \citet{2016MNRAS.457.3817S}, we see the contrast between the density peaks and the density minimum of a gap, as well as the width, grow in time, somewhere between $\propto \sqrt{t}$ and linear. In this simplified 1D collisionless dispersion analysis, the gaps between the different packets space apart linearly in time with $\delta x\propto \delta vt$, where $\delta v$ is the mean velocity of the about the impact site for two different packets. This inclines us to believe that the gaps space apart before they become significant. }

\salvatore{Nonetheless, we believe this is the first consideration of decomposing the stream into packets of tidal bursts to understand the effect of the internal dynamics on gap persistence. Since gap persistence depends on the mass and internal evolution of the cluster, characterizing the gap can constrain our knowledge of the internal evolution. }
% \paola{What happens is that the different and in particular in the populations of stars in this part of the tail that escaped from the cluster at different times, in each of these populations the respective gap tends to travel along the streams at different speeds, which leads to a phase shift of the gap and its loss of visibility. In other words, the gap can survive for billions of years in each population of stars that left the cluster at different times, but it is no longer. Still, initially coherent and therefore visible because each of these populations travels along the tail (and drags a gap with it) at different speeds. The loss of coherence occurs at a time $t_{visibility}$ which is directly proportional to the gap size and inversely proportional to the velocity dispersion in that part of the stream. So, if populations of stars lost at different times co-exist in the portion of the stream where the impact occurs, the gap will remain visible for a time that depends on the mass of the progenitor system at the time these stars escaped. From this, we can understand that various factors determine the visibility of a gap in a stream. Mass affects the velocity dispersion in a given region of the stream, but this velocity dispersion, in turn, depends on the history of mass loss of the cluster. Suppose stars dominate a given stream region, and all are lost simultaneously from the cluster. In that case, the resulting velocity dispersion is lower than if various populations coexist in the same region, which has implications for these structures' persistence. }

\begin{figure*}
    \includegraphics[width=\linewidth]{decoherence_shocked_outflow.png}
    \caption{We dissect the density of the stream from the bottom panel of Fig.~\ref{fig:gap_no_more} into stars that are either shocked or leaked out, see text for details. The bottom panel shows the total 1D profile with the two different populations. The second from the bottom dissects the total shocked outflow into individual pericenter passages. Each of which is color-coated by the escape time. The top two panels separate the first four shocked outflow populations to aid visualization of the gap placement from the impact of NGC~7078. }
    \label{fig:decoherence_shocked_outflow}
\end{figure*}
